The role of cilia-mediated cerebrospinal fluid flow in intraventricular transport in zebrafish embryos

### 2025-02-03 paper/main.tex
Edit in Methods - Part III

\subsection*{Computational model of solute transport after photoconversion}
We simulate transport of photoconverted proteins within the brain ventricles by modeling transport of a solute with concentration $c(\xx, t)$ for $\xx\in\Omega$ and time $t > 0$ by the advection-diffusion equation
\begin{equation}
    \pdifft{c} + \nabla\cdot\JJ = 0 \quad \text{in } \Omega,
    \label{eq:adv_diff_strong}
\end{equation}
where the total concentration flux $\mathbf{J}$ is assumed to consist of an advective and a diffusive flux:
\begin{equation*}
    \JJ = c\uu - D\nabla c .
\end{equation*}
Here, $\uu$ is the cerebrospinal fluid velocity field, governed by the Stokes equations~\eqref{eq:stokes_eqs}, and $D$ is the molecular diffusion coefficient of the solute. Initially, we set $c(\xx, 0)=0$ for all $\xx\in\Omega$. The ventricular walls are assumed to be impermeable (admitting \emph{no flux} i.e.~$\JJ\cdot\nn=0$), except at the anterior/posterior boundary $\Gamma_p$. \fixme{There, we consider either prescribed \emph{influx} in the form of a time-periodic condition on the total flux $(\JJ\cdot\nn)|_{\mathrm{in}}^{n+1}=(\JJ\cdot\nn)|_{\mathrm{out}}^{n}$ and/or \emph{outflux} with zero diffusive flux $D\nabla c\cdot\nn=0$.}
\mer{MER: Could we make this section about the boundary conditions easier to read? I've had a go at it, please check.}

@@

\subsection*{Physical parameters}
%\label{subsec:parameters}
The cerebrospinal fluid density $\rho$ and dynamic viscosity $\mu$ were set following Bloomfield \textit{et al.}~\cite{Bloomfield1998EffectsFluid}~(\cref{tab:parameter_table}). We considered small, medium and large-sized molecules by three different diffusion coefficients $D_1, D_2$ and $D_3$ (\Cref{tab:parameter_table}). Experimental data~\cite{Olstad2019CiliaryDevelopment} indicate \fixme{velocities} proximate to the ventricular walls of $27.4 \pm 5.4$ \textmu m/s and $2.6 \pm 0.6$ \textmu m/s in the dorsal and ventral regions of the diencephalic ventricle, respectively. In the telencephalic ventricle, the velocities were estimated at $4.7 \pm 1.3$ \textmu m/s. We calibrated $\tau$ and $A$ heuristically by comparing these experimental velocity values with the simulated, maximum velocity magnitude. In addition, particle tracking data of pulsatile motion~(\cref{sec:appC_particle_transport}) was used in the calibration process to determine the relative contributions of the cardiac pulsations, as compared to the cilia-driven flow, to the velocity magnitude.
\mer{MER: What velocities? Net velocities? Velocity magnitude aka speed? Also, haven't you said this part about the experimental data already? Either place the details there, or all here. I suggest there.}
\mer{MER: I suggest you move the parts relevant for cilia/cardiac flow to where you mentioned it first, and then merge this section with the Estimation of diffusion .... Just refer to \cref{tab:parameter_table}. Actually, unsure if you need it. You say everything in the text or in the other Table right?}
\begin{table}[h!]
    \centering
    \begin{tabular}{lcccc}
    \toprule
        Parameter & Symbol & Value & Unit & Reference \\
        \midrule
        Fluid density & $\rho$ & $1000$ & $\mathrm{kg/m^3}$ & \cite{Bloomfield1998EffectsFluid} \\
        Dynamic viscosity & $\mu$ & $\num{6.97e-04}$ & $\mathrm{Pa\cdot s}$ & \cite{Bloomfield1998EffectsFluid} \\
        Cardiac frequency & $f$ & 2.22 & Hz &  \cite{Olstad2019CiliaryDevelopment}\\
        Angular cardiac frequency & $\omega = 2\pi f$ & 6.97 & rad/s & \cite{Olstad2019CiliaryDevelopment}\\
        Cilia tangential traction & $\tau$ & \num{6.5e-04} & Pa & * \\
        Pressure amplitude & $A$ & \num{1.5e-03} & Pa & * \\
        %Ref. frame motion amplitude & $A_{\mathrm{ref}}$ & \num{9.19e-04} & m/s & *\\
        Exosome diffusion coefficient & $D_1$ & $\num{1.63e-12}$ & $\mathrm{m^2/s}$ & \textdagger\\
        STM+GFP diffusion coefficient  & $D_2$ & $\num{5.75e-11}$ & $\mathrm{m^2/s}$ & \textdagger\\
        Dendra2 diffusion coefficient  & $D_3$ & $\num{1.15e-10}$ & $\mathrm{m^2/s}$ & \textdagger\\
    \bottomrule
    \end{tabular}
    \caption{Physical parameters of the flow and transport models. (*) Calibrated based on experimental data~\cite{Olstad2019CiliaryDevelopment}. (\textdagger) Estimated based on experimental data~\cite{GuraSadovsky2017MeasurementExpansion, Swaminathan1997PhotobleachingDiffusion, Potma2001ReducedCells}.}
    \label{tab:parameter_table}
\end{table}

@@

\mer{MER: Let's discuss the paragraphs below and see where they fit best.}
When simulating transport of solute in the ventricles, we consider the same six different regions of interest that were considered in the photoconversion experiments. In region number one (the photoconversion region), we use the function~\eqref{eq:injection_curve} to enforce the concentration such that it represents the experimental data of photoconversion of proteins. We consider the three different diffusion coefficients introduced in previous sections to simulate and compare results for transport of molecules of varying sizes.

@@

\subsection*{Quantities of interest}
We quantify the transport of solute by calculating the mean concentrations
\begin{equation}
    \overline{c}_i(t) = \frac{1}{|\Omega_i|}\int_{\Omega_i} c(\xx, t)\,\dx,
    \label{eq:c_mean_i}
\end{equation}
in the regions of interest (ROIs) $\Omega_i$ as functions of time, where $|\Omega_i|$ is the volume of the domain $\Omega_i$. The dynamics of ROI 1 follows from imposing~\cref{eq:injection_curve}. As another way of quantifying transport, we report the times $\hat{t}_i$ when the mean concentrations $\overline{c}_i$ first exceed a threshold value $\hat{c}_i$. We use $\hat{c}_i=0.25$ for regions of interest 1--4 and $\hat{c}_i=0.10$ for regions 5 and 6.

To evaluate the relative importance of advection and diffusion as transport mechanisms, we consider the Péclet number
\begin{equation*}
    \mathrm{Pe} = \frac{\mathrm{diffusion \ timescale}}{\mathrm{advection \ timescale}} = \frac{t_d}{t_a} = \frac{L_p^2/D}{L_p/U_p} = \frac{L_p U_p}{D}.
\end{equation*}
Here, $L_p$ and $U_p$ are characteristic length and velocity scales, respectively, whereas $D$ is the solute diffusion coefficient as previously. To assess the assumption that the CSF flow is governed by the Stokes equations, we also compute the Reynolds number
\begin{equation*}
    \mathrm{Re} = \frac{\mathrm{inertial \ forces}}{\mathrm{viscous \ forces}} = \frac{\rho U_r L_r}{\mu},
\end{equation*}
with $L_r$ and $U_r$ as the length and velocity scales characteristic of the flow.

%-----DISCRETIZATION AND MODEL IMPLEMENTATION-----%
\subsection*{Numerical approximation of the flow equations}

\mer{associated with a total of 943 820 degrees of freedom}

The Stokes equations~\eqref{eq:stokes_eqs} were discretized with a finite element method. In time, we employ a first-order implicit Euler discretization. We chose the timestep size $\Delta t = 0.02252$ s such that 20 timesteps represent one cardiac cycle. For the spatial discretization, we used piecewise linear Brezzi-Douglas-Marini (BDM) elements for the velocity. The BDM elements have degrees of freedom associated with integral moments of the velocity component normal to the facets of elements~\cite{Brezzi1985TwoProblems}, thus we enforced the impermeability condition $\uu\cdot\nn=0$ strongly. We imposed the boundary conditions for both the tangential traction $\bsigpar$ and the normal pressure $\Tilde{p}(t)$ weakly. For the pressure, we used piecewise constant elements, referred to as zeroth order discontinuous Galerkin (DG) elements. The element combination $\mathrm{BDM}_1$--$\mathrm{DG}_{0}$ yields a stable discretization scheme for the Stokes equations~\cite{Stenberg1989SomeEquations}. Since the divergence of the velocity is a subspace of the pressure space for this scheme, the divergence-free condition of the velocity is satisfied exactly on each element~\cite{Boffi2008FiniteProblem}, ensuring mass conservation. The BDM discretization includes a penalty parameter $\gamma$ which needs to be chosen large enough to ensure stability of the finite element method. We set $\gamma=10.$

In the absence of pressure boundary conditions, the pressure in the Stokes equations is only determined up to an additive constant. Thus, with the boundary conditions used in model version I, the pressure is not uniquely defined and discretization of the Stokes equations leads to a singular linear equation system. To make the discrete problem well-posed for this setup, the right-hand side vector of the linear system assembled after discretization of the Stokes equations equations is orthogonalized with respect to the basis of the nullspace of the pressure to ensure existence of solutions. This nullspace is the set of (real) scalar constants. Furthermore, information of the pressure nullspace is provided to the linear solver, such that a unique solution is determined.

\subsection*{Numerical approximation of the transport equations}
The advection-diffusion equation~\eqref{eq:adv_diff_strong} is discretized with a first-order implicit Euler method in time, and a first-order, symmetric interior penalty discontinuous Galerkin (DG) method in space~\cite{Arnold1982AnElements}. Like for the flow equations, we use a timestep size of $\Delta t = 0.02252$ s. We simulate transport for 1900 cardiac cycles, meaning the final time of simulation $T = 856$ s. The DG penalty parameter $\alpha = 25$ was chosen large enough to ensure stability of the finite element method. To accurately approximate the solute concentrations even in the presence of strong advection (high Péclet numbers), we use an upwind scheme for the velocity when calculating the advective flux~\cite{Patankar2018NumericalFlow}.

\subsection*{Solution strategy and software implementation}
Because of the one-way coupling between the velocity field $\uu$ and the concentration $c$, the governing equations for the cerebrospinal fluid flow and the concentration can be solved sequentially when modeling solute transport. For model version I, the flow field is a steady-state solution. In this case, we first solve the Stokes equations~\eqref{eq:stokes_eqs} for $\uu$ and $p$ until we reach the steady-state solution, and then use the velocity field $\uu$ as input when solving the advection-diffusion equation~\eqref{eq:adv_diff_strong}. For the baseline model 0 and model II, the Stokes equations are solved for one period of the pulsatile cardiac motion. The solution of this one period is then used periodically as input when solving the advection-diffusion equation.

The finite element methods derived for the Stokes equations and the advection-diffusion equation were implemented and solved numerically with DOLFINx, the Python interface of the finite element software FEniCSx~\cite{TheFEniCSProject2024FEniCSxDocumentation}, with petsc4py~\cite{Dalcin2011ParallelPython} as linear algebra backend. For solving the linear systems, a direct solver using MUMPS~\cite{Amestoy2011Mumps} was used to solve the discretized Stokes equations. For the discretized advection-diffusion equation, we used an iterative solver based on the GMRES routine~\cite{Saad1986GMRES:Systems} with block Jacobi preconditioning~\cite{Jacobi1845UeberGleichungen}. Numerical experiments verified the accuracy of the model implementation (\cref{sec:appB_verification}). The implementation of the model presented in this work is available at~\url{insert-repo-link-here}.

>>>>

\subsection*{Computational model of solute transport after photoconversion}
We simulate transport of photoconverted proteins within the brain ventricles by modeling transport of a solute with concentration $c(\xx, t)$ for $\xx\in\Omega$ and time $t > 0$ by the advection-diffusion equation
\begin{equation}
    \pdifft{c} + \nabla\cdot\JJ = 0 \quad \text{in } \Omega,
    \label{eq:adv_diff_strong}
\end{equation}
where the total concentration flux $\mathbf{J}$ is assumed to consist of an advective and a diffusive flux:
\begin{equation*}
    \JJ = c\uu - D\nabla c .
\end{equation*}
Here, $\uu$ is the CSF velocity field, governed by the Stokes equations~\eqref{eq:stokes_eqs}, and $D$ is the molecular diffusion coefficient of the solute. Initially, we set $c(\xx, 0)=0$ for all $\xx\in\Omega$. The ventricular walls are assumed to be impermeable (admitting \emph{no flux} i.e.~$\JJ\cdot\nn=0$), except at the anterior/posterior boundary $\Gamma_p$. \fixme{There, we consider either prescribed \emph{influx} in the form of a time-periodic condition on the total flux $(\JJ\cdot\nn)|_{\mathrm{in}}^{n+1}=(\JJ\cdot\nn)|_{\mathrm{out}}^{n}$ and/or \emph{outflux} with zero diffusive flux $D\nabla c\cdot\nn=0$.}
\mer{MER: Could we make this section about the boundary conditions easier to read? I've had a go at it, please check.}

@@

\subsection*{Physical parameters}
%\label{subsec:parameters}
The CSF density $\rho$ and dynamic viscosity $\mu$ were set following Bloomfield \textit{et al.}~\cite{Bloomfield1998EffectsFluid}~(\cref{tab:parameter_table}). We considered small, medium and large-sized molecules by three different diffusion coefficients $D_1, D_2$ and $D_3$ (\Cref{tab:parameter_table}). Experimental data~\cite{Olstad2019CiliaryDevelopment} indicate \fixme{velocities} proximate to the ventricular walls of $27.4 \pm 5.4$ \textmu m/s and $2.6 \pm 0.6$ \textmu m/s in the dorsal and ventral regions of the diencephalic ventricle, respectively. In the telencephalic ventricle, the velocities were estimated at $4.7 \pm 1.3$ \textmu m/s. We calibrated $\tau$ and $A$ heuristically by comparing these experimental velocity values with the simulated, maximum velocity magnitude. In addition, particle tracking data of pulsatile motion~(\cref{sec:appC_particle_transport}) was used in the calibration process to determine the relative contributions of the cardiac pulsations, as compared to the cilia-driven flow, to the velocity magnitude.
\mer{MER: What velocities? Net velocities? Velocity magnitude aka speed? Also, haven't you said this part about the experimental data already? Either place the details there, or all here. I suggest there.}
\mer{MER: I suggest you move the parts relevant for cilia/cardiac flow to where you mentioned it first, and then merge this section with the Estimation of diffusion .... Just refer to \cref{tab:parameter_table}. Actually, unsure if you need it. You say everything in the text or in the other Table right?}
\begin{table}[h!]
    \centering
    \begin{tabular}{lcccc}
    \toprule
        Parameter & Symbol & Value & Unit & Reference \\
        \midrule
        Fluid density & $\rho$ & $1000$ & $\mathrm{kg/m^3}$ & \cite{Bloomfield1998EffectsFluid} \\
        Dynamic viscosity & $\mu$ & $\num{6.97e-04}$ & $\mathrm{Pa\cdot s}$ & \cite{Bloomfield1998EffectsFluid} \\
        Cardiac frequency & $f$ & 2.22 & Hz &  \cite{Olstad2019CiliaryDevelopment}\\
        Angular cardiac frequency & $\omega = 2\pi f$ & 6.97 & rad/s & \cite{Olstad2019CiliaryDevelopment}\\
        Cilia tangential traction & $\tau$ & \num{6.5e-04} & Pa & * \\
        Pressure amplitude & $A$ & \num{1.5e-03} & Pa & * \\
        %Ref. frame motion amplitude & $A_{\mathrm{ref}}$ & \num{9.19e-04} & m/s & *\\
        Exosome diffusion coefficient & $D_1$ & $\num{1.63e-12}$ & $\mathrm{m^2/s}$ & \textdagger\\
        STM+GFP diffusion coefficient  & $D_2$ & $\num{5.75e-11}$ & $\mathrm{m^2/s}$ & \textdagger\\
        Dendra2 diffusion coefficient  & $D_3$ & $\num{1.15e-10}$ & $\mathrm{m^2/s}$ & \textdagger\\
    \bottomrule
    \end{tabular}
    \caption{Physical parameters of the flow and transport models. (*) Calibrated based on experimental data~\cite{Olstad2019CiliaryDevelopment}. (\textdagger) Estimated based on experimental data~\cite{GuraSadovsky2017MeasurementExpansion, Swaminathan1997PhotobleachingDiffusion, Potma2001ReducedCells}.}
    \label{tab:parameter_table}
\end{table}

@@

\mer{MER: Let's discuss the paragraphs below and see where they fit best. I think we can drop the first paragraph (you said this already). I find the second paragraph to be somewhat confusing (low signal-to-noise ratio).}
When simulating transport of solute in the ventricles, we consider the same six different regions of interest that were considered in the photoconversion experiments. In region number one (the photoconversion region), we use the function~\eqref{eq:injection_curve} to enforce the concentration such that it represents the experimental data of photoconversion of proteins. We consider the three different diffusion coefficients introduced in previous sections to simulate and compare results for transport of molecules of varying sizes.

@@

\subsection*{Quantities of interest}
We calculate the mean solute concentrations
\begin{equation}
    \overline{c}_i(t) = \frac{1}{|\Omega_i|}\int_{\Omega_i} c(\xx, t)\,\dx,
    \label{eq:c_mean_i}
\end{equation}
in the six regions of interest (ROIs) $\Omega_i$ ($i = 1, \dots, 6$) as functions of time, where $|\Omega_i|$ is the volume of the domain $\Omega_i$. The dynamics of ROI 1 follows directly from~\eqref{eq:injection_curve}. In addition, we report time-to-threshold: the time $\hat{t}_i$ when the mean concentration $\overline{c}_i$ first exceed a threshold value $\hat{c}_i$. We use $\hat{c}_i=0.25$ for ROI 1--4 and $\hat{c}_i=0.10$ for ROI 5--6.

To evaluate the relative importance of advection and diffusion as transport mechanisms, we consider the Péclet number
\begin{equation*}
    \mathrm{Pe} = \frac{\mathrm{diffusion \ timescale}}{\mathrm{advection \ timescale}} = \frac{t_d}{t_a} = \frac{L_p^2/D}{L_p/U_p} = \frac{L_p U_p}{D}.
\end{equation*}
Here, $L_p$ and $U_p$ are characteristic length and velocity scales, respectively, and $D$ is the solute diffusion coefficient. To assess the assumption that the CSF flow is governed by the Stokes equations, we also compute the Reynolds number
\begin{equation*}
    \mathrm{Re} = \frac{\mathrm{inertial \ forces}}{\mathrm{viscous \ forces}} = \frac{\rho U_r L_r}{\mu},
\end{equation*}
with $L_r$ and $U_r$ as length and velocity scales characteristic of the flow.

%-----DISCRETIZATION AND MODEL IMPLEMENTATION-----%
\subsection*{Numerical approximation of the Stokes equations}

The Stokes equations~\eqref{eq:stokes_eqs} are discretized with a finite element method in space, while in time, we employ a first-order implicit Euler discretization. We set the timestep size $\Delta t = 0.02252$ s such that 20 timesteps represent one cardiac cycle. For the spatial discretization, we use piecewise linear Brezzi-Douglas-Marini (BDM) elements for the velocity. The BDM elements have degrees of freedom associated with integral moments of the velocity component normal to the facets of elements~\cite{Brezzi1985TwoProblems}, thus we enforce the impermeability condition $\uu\cdot\nn=0$ strongly. We impose the boundary conditions for both the tangential traction $\bsigpar$ and the normal pressure $\Tilde{p}(t)$ weakly. For the pressure, we use piecewise constant (zeroth order discontinuous Galerkin (DG)) elements. The resulting total number of degrees of freedom at each time step is 943 820. \fixme{The element combination $\mathrm{BDM}_1$--$\mathrm{DG}_{0}$ yields a non-conforming but stable discretization scheme for the Stokes equations~\cite{Stenberg1989SomeEquations}. The discretization includes a penalty parameter $\gamma$ which needs to be chosen large enough to ensure stability of the finite element method. We set $\gamma=10$.} Since the divergence of the velocity is a subspace of the pressure space for this scheme, the divergence-free condition of the velocity is satisfied exactly on each element~\cite{Boffi2008FiniteProblem}, ensuring mass conservation.

\mer{MER: I think this is too much detail. One sentence suffices, or it goes in the Supplementary.}
In the absence of pressure boundary conditions, the pressure in the Stokes equations is only determined up to an additive constant. Thus, with the boundary conditions used in model version I, the pressure is not uniquely defined and discretization of the Stokes equations leads to a singular linear equation system. To make the discrete problem well-posed for this setup, the right-hand side vector of the linear system assembled after discretization of the Stokes equations equations is orthogonalized with respect to the basis of the nullspace of the pressure to ensure existence of solutions. This nullspace is the set of (real) scalar constants. Furthermore, information of the pressure nullspace is provided to the linear solver, such that a unique solution is determined.

\subsection*{Numerical approximation of the transport equations}
The advection-diffusion equation~\eqref{eq:adv_diff_strong} is discretized with a first-order implicit Euler method in time, and a first-order, symmetric interior penalty discontinuous Galerkin (DG) method in space~\cite{Arnold1982AnElements}. To accurately approximate the solute concentrations even in the presence of strong advection (high Péclet numbers), we use an upwind scheme for the velocity when calculating the advective flux~\cite{Patankar2018NumericalFlow}. As for the flow equations, we use a timestep size of $\Delta t = 0.02252$ s. We simulate transport for 1900 cardiac cycles, with a final time of simulation $T = 856$ s. A DG penalty parameter $\alpha = 25$ was chosen to ensure stability of the method.

\subsection*{Solution strategy and software implementation}
Because of the one-way coupling between the velocity field $\uu$ and the concentration $c$, the governing equations for the CSF flow and the concentration can be solved sequentially when modeling solute transport. \fixme{For the no-cardiac/cilia-only flow scenario, the velocity quickly reaches steady-state.} In this case, we first solve the \fixme{steady-state} Stokes equations~\eqref{eq:stokes_eqs} for $\uu$ and $p$, and then use the velocity field $\uu$ as input when solving the advection-diffusion equation~\eqref{eq:adv_diff_strong}. For the other flow scenarios, the Stokes equations are solved for one period of the pulsatile cardiac motion. The solution of this one period is then used periodically as input for the advection-diffusion equation.

The numerical solution schemes for the Stokes equations and the advection-diffusion equation were implemented and solved numerically with DOLFINx, the Python interface of the finite element software FEniCSx~\cite{TheFEniCSProject2024FEniCSxDocumentation}, with petsc4py~\cite{Dalcin2011ParallelPython} as linear algebra backend. A direct solver using MUMPS~\cite{Amestoy2011Mumps} was used to solve the linear systems resulting from the discretized Stokes equations. For the discretized advection-diffusion equation, we used an iterative linear solver based on GMRES~\cite{Saad1986GMRES:Systems} with block Jacobi preconditioning~\cite{Jacobi1845UeberGleichungen}. Numerical experiments verified the accuracy of the model implementation (\cref{sec:appB_verification}). This simulation software is openly available~\cite{zenodo-link-to-all-code}.
